---
title: "Bayesian Comparison Of Home Run Data (2022 and 2001)"
author: "Aidan Cooney"
date: "2023-04-19"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(rvest)
library(invgamma)

# code to create the normal-normal-inverse gamma gibbs sampler
nnig.mcmc <- function(data, m, v, a, b, J) {
  
  ybar <- mean(data)
  s2 <- var(data)
  n <- length(data)

  # create empty vectors 
  mu <- numeric()
  sigma2 <- numeric()
  
  # initial values
  mu[1] <- ybar
  sigma2[1] <- s2

  J <- 10000

  for (j in 2:J){
    # update mu
    mstar <- (n * v * ybar + m * sigma2[j - 1]) / (n * v + sigma2[j - 1])
    vstar <- (v * sigma2[j - 1]) / (n * v + sigma2[j - 1])
    mu[j] <- rnorm(1, mstar, sqrt(vstar))
  
    # update sigma2
    astar <- a + 0.5 * n
    bstar <- b + 0.5 * sum((data - mu[j])^2)
    sigma2[j] <- rinvgamma(1, astar, bstar)
  }

  # output the gibbs sampler draws 
  out <- cbind(mu, sigma2)
  colnames(out) <- c("mu", "sigma2")
  out
}

```

```{r, include = FALSE}
set.seed(101)
```

```{r, include = FALSE}
mlb.2022 <- "https://www.baseball-reference.com/leagues/majors/2022.shtml" %>%
  read_html() %>%
  html_node("table") %>%
  html_table() %>%
  as_tibble()

mlb.2022 <- mlb.2022[-c(31, 32, 33), ]
mlb.2022 <- mlb.2022[["HR"]]
mlb.2022 <- as.numeric(mlb.2022)

mlb.2001 <- "https://www.baseball-reference.com/leagues/majors/2001.shtml" %>%
  read_html() %>%
  html_node("table") %>%
  html_table() %>%
  as_tibble()

mlb.2001 <- mlb.2001[-c(31, 32, 33), ]
mlb.2001 <- mlb.2001[["HR"]]
mlb.2001 <- as.numeric(mlb.2001)
```


```{r, include = FALSE}
m <- 150
v <- 400
a <- 3
b <- 3200
J <- 10000
```


## Introduction 

Major league baseball has seen home run numbers rise over the past few seasons, 
showing that teams appear to be placing an emphasis on players that can hit the 
ball out of the ballpark. To truly test how high the home run numbers have gotten, 
I chose to test data from last season (2022) to the home run numbers from a season
during the height of the steroid era of baseball (2001). The steroid era was 
between 1994 and 2004, during which numerous players admitted to and tested 
positive for performance enhancing drugs (PEDS) such as steroids. Through this 
analysis, I hope to determine if there is a significant difference between the 
home run numbers in 2022 and the home run numbers in 2001. This 
analysis could show just how much emphasis teams are currently placing on home runs 
as they once did during the steroid era. 

## Methods

For this analysis, I found the data on a website called "baseball reference", 
which contains data collected from each mlb season. I filtered mlb 
seasons to find the information from 2022, and the information from 2001 to 
perform my analysis. 

The data model I am using to represent my data in this analysis is a normal-normal-inverse 
gamma distribution, as both the population mean and variance are 
unknown parameters. Since both are unknown, using a Gibbs Sampler with the 
normal-normal-inverse gamma is a reasonable distribution as we would be able to 
estimate the unknown parameters in our model with reasonable accuracy. 
The parameters of interest for the likelihood would be the 
mean and variance for each year.

For my prior belief for $\mu$ I chose a normal distribution with a mean value of 
150 and a variance of 400 ($\mu \sim N(m = 150, v = 400)$). This is a reasonable 
prior distribution as this would put the expected value of the prior 
to be 150 home runs, which appears to be a reasonable prior average per team. 

For my prior belief for $\sigma^2$ I chose an inverse gamma with a shape of 3 and 
a rate of 3200 ($\sigma^2 \sim IG(a = 3, b = 3200)$). This is a reasonable prior 
as this would put the expected value for the variance at 1600, making the expected 
standard deviation around 40 home runs which seems to be a reasonable estimate 
for the variance. Combining this expected value with that of the expectation for 
the mean, an interval that is one standard deviation from the mean would be between
110 to 190 which is where most of the data should be contained. This appears to 
be a reasonable interval before seeing any data. 


## Results

### Posterior For 2022

```{r, include = FALSE}
post.2022 <- nnig.mcmc(mlb.2022, m, v, a, b, J)
```

```{r, include = FALSE}
# code to test for convergence without any burn in 
plot(post.2022[, 1], type = 'l')
plot(post.2022[, 2], type = 'l')
acf(post.2022[,1])
acf(post.2022[,2])

keep <- 100:J

# code to test for convergence with burn in included 
plot(post.2022[keep, 1], type = 'l')
plot(post.2022[keep, 2], type = 'l')
acf(post.2022[keep, 1])
acf(post.2022[keep, 2])
```

```{r, include=FALSE}
mean(post.2022[keep, 1])
mean(post.2022[keep, 2])

quantile(post.2022[keep,1], c(0.025, 0.975))
quantile(post.2022[keep, 2], c(0.025, 0.975))
```

After performing a Gibbs sampler to draw from the posterior for 2022, the expected 
value for the mean of our posterior given our prior knowledge and accounting for 
the data was around 171.62 with an expected variance of around 1262.34, which 
would make the expected standard deviation around 35.53. 

Looking at the credible interval for the mean, there is a 95% probability that 
the average number of home runs is between 159.43 and 183.47 home runs given our 
prior knowledge and accounting for the data. 

Looking next at the credible interval for the variance of the posterior, there is 
a 95% probability that the variance is between 784.42 and 2013.03 given our prior 
knowledge and accounting for the data. 


```{r, echo = FALSE}
vals <- seq(90, 230, length = 1001)
plot(density(post.2022[keep, 1]), xlim = c(90, 230), type = 'l', lwd = 3, 
     main = "Prior and Posterior for Mean (2022)", xlab = expression(mu))
lines(vals, dnorm(vals, m, sqrt(v)), type = 'l', col = 'red', lwd = 3)
legend(x = 'topright', legend = c('Prior', 'Posterior'), col = c('red', 'black'),
       lty = 1, lwd = 3)
```

Looking at the prior and posterior plot for the average number of home runs for 
2022, we can see that the posterior plot is far more peaked with less variability 
than the prior plot. The posterior distribution is centered around 170 home runs after 
updating for the data gathered, which is larger than the prior's center of 150 home runs. 


```{r, echo = FALSE}
sig2_vec <- seq(0, 15000, length=1001)
plot(density(post.2022[keep, 2]), type = 'l', xlim = c(0, 3500), lwd = 3, 
     main = "Prior and Posterior for sigma2 (2022)", xlab = expression(sigma^2))
lines(sig2_vec, dinvgamma(sig2_vec, a, b), type = 'l', lwd = 3, col = 'red')
legend(x = 'topright', legend = c('Prior', 'Posterior'), col = c('red', 'black'),
       lty = 1, lwd = 3)
```

Looking at the plots for the prior and posterior distributions for $\sigma^2$ we 
can see that after updating for the data, the center of the variance has shifted 
to the right and has a higher peak, with less variability. The expected variance 
for the posterior is larger than the expected variance for the prior for 2022. 

### Posterior For 2001

```{r, include = FALSE}
post.2001 <- nnig.mcmc(mlb.2001, m, v, a, b, J)
```

```{r, include = FALSE}
# code to test for convergence without any burn in 
plot(post.2001[, 1], type = 'l')
plot(post.2001[, 2], type = 'l')
acf(post.2001[,1])
acf(post.2001[,2])

keep <- 100:J

# code to test for convergence with burn in included 
plot(post.2001[keep, 1], type = 'l')
plot(post.2001[keep, 2], type = 'l')
acf(post.2001[keep, 1])
acf(post.2001[keep, 2])
```

```{r, include = FALSE}
mean(post.2001[keep, 1])
mean(post.2001[keep, 2])

quantile(post.2001[keep, 1], c(0.025, 0.975))
quantile(post.2001[keep, 2], c(0.025, 0.975))
```

After performing a Gibbs sampler to draw from the posterior for 2001, the expected 
value for the mean of our posterior given our prior knowledge and accounting for 
the data was around 179.31 with an expected variance of around 1081.22, which 
would make the expected standard deviation around 32.88. 

Looking at the credible interval for the mean, there is a 95% probability that 
the average number of home runs is between 167.93 and 190.55 home runs given our 
prior knowledge and accounting for the data. 

Looking next at the credible interval for the variance of the posterior, there is 
a 95% probability that the variance is between 668.05 and 1734.32 given our prior 
knowledge and accounting for the data. 

```{r, echo = FALSE}
vals <- seq(90, 230, length = 1001)
plot(density(post.2001[keep, 1]), xlim = c(90, 230), type = 'l', lwd = 3, 
     main = "Prior and Posterior for Mean (2001)", xlab = expression(mu))
lines(vals, dnorm(vals, m, sqrt(v)), type = 'l', col = 'red', lwd = 3)
legend(x = 'topright', legend = c('Prior', 'Posterior'), col = c('red', 'black'),
       lty = 1, lwd = 3)
```

Looking at the plot for the posterior and prior of the mean for 2001, we can see 
that the posterior distribution has a much higher peak and less variability than 
the prior distribution. The posterior for the mean is also shifted to the right, 
as the center of the posterior distribution is around 180 while the prior center 
is at 150. 

```{r, echo = FALSE}
sig2_vec <- seq(0, 15000, length=1001)
plot(density(post.2001[keep, 2]), type = 'l', xlim = c(0, 3500), lwd = 3, 
     main = "Prior and Posterior for sigma2 (2001)", xlab = expression(sigma^2))
lines(sig2_vec, dinvgamma(sig2_vec, a, b), type = 'l', lwd = 3, col = 'red')
legend(x = 'topright', legend = c('Prior', 'Posterior'), col = c('red', 'black'),
       lty = 1, lwd = 3)
```

Looking at the prior and posterior distributions for the variance of 2001, we can see 
that the center of the posterior distribution appears to be shifted slightly to 
the right, and has a higher peak than the prior distribution. The posterior 
distribution also appears to have less variablity than the prior distribution. 


### Posterior Inference For The Difference (2022 - 2001)

```{r, echo = FALSE}
plot(density(post.2022[keep, 1] - post.2001[keep, 1]), type = 'l',
     main = "Difference in Posterior Mean (2022 - 2001)", xlab = "Difference in Mean")
```

Looking at the plot for the difference in the mean of the posterior distributions 
for 2022 and 2001, we can see that the center of this plot appears to be around 
-8, meaning we expect, on average, more home runs for each team in 2001 
compared to 2022. However, 0 does appear to be within the distribution, so a look at a 
credible interval can help show if the difference is truly significant. 

```{r, include = FALSE}
mean(post.2022[keep, 1] - post.2001[keep, 1] < 0)

quantile(post.2022[keep, 1] - post.2001[keep, 1], c(0.025, 0.975))
```

Looking at the difference in mean home runs (2022 - 2001) the probability that the difference 
is less than 0 is around 82.35%, meaning that 2001 tended to experience more 
home runs per team than 2022. Looking at the posterior credible interval however, we get 
an interval of -23.95 to 8.39 home runs. Since 0 is contained within this interval 
the difference in means does not appear to be significant. 

```{r, echo = FALSE}
plot(density(post.2022[keep, 2] / post.2001[keep, 2]), type = 'l',
     main = "Ratio of Variances (2022 / 2001)", xlab = expression(sigma^2))
```

Looking at the plot for the ratio of variances (2022 / 2001), the plot appears to 
be centered slightly above 1, meaning that we would expect the variance for 2022 
to be slightly larger than 2001 However 1 is contained in the distribution so a look at 
a credible interval can show significance. 

```{r, include = FALSE}
mean((post.2022[keep, 2] / post.2001[keep, 2]) > 1)

quantile(post.2022[keep, 2] / post.2001[keep, 2], c(0.025, 0.975))
```

Looking at the ratio of variances for the posteriors, the probability that the ratio of 
variances is greater than 1 is about 68.20%, meaning that most of the 
variances for 2022 are larger than the variances for 2001. However, looking at the 
posterior credible interval for the ratio of variances, we received an interval of 
0.59 to 2.25. Since 1 is contained in the interval, there is potentially no 
significant difference in variances between 2022 and 2001. 


## Discussion

After performing a Bayesian analysis on the data from 2022 and 2001, there does 
not appear to be a significant difference in the means or variances for 2022 
compared to 2001. This would tell us that the home run numbers for these two years 
are fairly similar, with a fairly similar spread as well. 

The possible shortcomings for this study would be a limitation on the data. To 
truly test if home run numbers are the same today as they were during the steroid 
era there would need to be multiple years tested in order to get the full picture 
of how much the sport of baseball may or may not have changed. Another test that could help 
illustrate this idea more clearly would be to compare each years data between 
2001 and 2022. This would show us if there was a decrease in 
home runs, and where this potential spike in home runs started up again after 
the steroid era ended. 

Overall, the analysis performed gave results for no significant difference in 
the mean and variances for the populations of average number of home runs. The number of 
home runs each team hit in 2022 and 2001 were roughly the same, meaning that baseball 
has placed an emphasis on power once again. Home runs are of high value in this 
modern era of baseball, and this analysis has shown just how valued they are.

## CODE 

```{r, eval = FALSE}
# libraries used 
library(tidyverse)
library(rvest)
library(invgamma)

# set seed to ensure same rnorm output values while comparing 
set.seed(101)

# code to create the normal-normal-inverse gamma gibbs sampler
nnig.mcmc <- function(data, m, v, a, b, J) {
  
  ybar <- mean(data)
  s2 <- var(data)
  n <- length(data)

  # create empty vectors 
  mu <- numeric()
  sigma2 <- numeric()
  
  # initial values
  mu[1] <- ybar
  sigma2[1] <- s2

  J <- 10000

  for (j in 2:J){
    # update mu
    mstar <- (n * v * ybar + m * sigma2[j - 1]) / (n * v + sigma2[j - 1])
    vstar <- (v * sigma2[j - 1]) / (n * v + sigma2[j - 1])
    mu[j] <- rnorm(1, mstar, sqrt(vstar))
  
    # update sigma2
    astar <- a + 0.5 * n
    bstar <- b + 0.5 * sum((data - mu[j])^2)
    sigma2[j] <- rinvgamma(1, astar, bstar)
  }

  # output the gibbs sampler draws 
  out <- cbind(mu, sigma2)
  colnames(out) <- c("mu", "sigma2")
  out
}

# code to read in 2022 dataset
mlb.2022 <- "https://www.baseball-reference.com/leagues/majors/2022.shtml" %>%
  read_html() %>%
  html_node("table") %>%
  html_table() %>%
  as_tibble()

# code to grab the necessary column of HR values
mlb.2022 <- mlb.2022[-c(31, 32, 33), ]
mlb.2022 <- mlb.2022[["HR"]]
# code to create this column as a numeric
mlb.2022 <- as.numeric(mlb.2022)

# code to grab the necessary column of HR values 
mlb.2001 <- "https://www.baseball-reference.com/leagues/majors/2001.shtml" %>%
  read_html() %>%
  html_node("table") %>%
  html_table() %>%
  as_tibble()

# code to grab the necessary columnn of HR values 
mlb.2001 <- mlb.2001[-c(31, 32, 33), ]
mlb.2001 <- mlb.2001[["HR"]]
# code to create this column as a numeric 
mlb.2001 <- as.numeric(mlb.2001)

# code to set up initial values 
m <- 150
v <- 400
a <- 3
b <- 3200
J <- 10000

# code to draw from the posterior of 2022
post.2022 <- nnig.mcmc(mlb.2022, m, v, a, b, J)

# code to test for convergence without any burn in 
plot(post.2022[, 1], type = 'l')
plot(post.2022[, 2], type = 'l')
acf(post.2022[,1])
acf(post.2022[,2])

keep <- 100:J

# code to test for convergence with burn in included 
plot(post.2022[keep, 1], type = 'l')
plot(post.2022[keep, 2], type = 'l')
acf(post.2022[keep, 1])
acf(post.2022[keep, 2])

# code to grab the expected mean and variance from the posterior 
mean(post.2022[keep, 1])
mean(post.2022[keep, 2])

# code to create 95% credible intervals for the mean and the variance of the posterior 
quantile(post.2022[keep,1], c(0.025, 0.975))
quantile(post.2022[keep,2], c(0.025, 0.975))

# code to plot the posterior mean for 2022
vals <- seq(90, 230, length = 1001)
plot(density(post.2022[keep, 1]), xlim = c(90, 230), type = 'l', lwd = 3, 
     main = "Prior and Posterior for Mean (2022)", xlab = expression(mu))
lines(vals, dnorm(vals, m, sqrt(v)), type = 'l', col = 'red', lwd = 3)
legend(x = 'topright', legend = c('Prior', 'Posterior'), col = c('red', 'black'),
       lty = 1, lwd = 3)

# code to plot the posterior variance for 2022
sig2_vec <- seq(0, 15000, length=1001)
plot(density(post.2022[keep, 2]), type = 'l', xlim = c(0, 3500), lwd = 3, 
     main = "Prior and Posterior for sigma2 (2022)", xlab = expression(sigma^2))
lines(sig2_vec, dinvgamma(sig2_vec, a, b), type = 'l', lwd = 3, col = 'red')
legend(x = 'topright', legend = c('Prior', 'Posterior'), col = c('red', 'black'),
       lty = 1, lwd = 3)

# code to draw from the posterior of 2001
post.2001 <- nnig.mcmc(mlb.2001, m, v, a, b, J)

# code to test for convergence without any burn in 
plot(post.2001[, 1], type = 'l')
plot(post.2001[, 2], type = 'l')
acf(post.2001[,1])
acf(post.2001[,2])

keep <- 20:J

# code to test for convergence with burn in included 
plot(post.2001[keep, 1], type = 'l')
plot(post.2001[keep, 2], type = 'l')
acf(post.2001[keep, 1])
acf(post.2001[keep, 2])

# code to grab the expected mean and variance for the posterior
mean(post.2001[keep, 1])
mean(post.2001[keep, 2])

# code to create 95% credible intervals for the mean and variance of the posterior
quantile(post.2001[keep, 1], c(0.025, 0.975))
quantile(post.2001[keep, 2], c(0.025, 0.975))

# code to plot the mean of the posterior for 2001
vals <- seq(90, 230, length = 1001)
plot(density(post.2001[keep, 1]), xlim = c(90, 230), type = 'l', lwd = 3, 
     main = "Prior and Posterior for Mean (2001)", xlab = expression(mu))
lines(vals, dnorm(vals, m, sqrt(v)), type = 'l', col = 'red', lwd = 3)
legend(x = 'topright', legend = c('Prior', 'Posterior'), col = c('red', 'black'),
       lty = 1, lwd = 3)

# code to plot the variance of the posterior for 2001
sig2_vec <- seq(0, 15000, length=1001)
plot(density(post.2001[keep, 2]), type = 'l', xlim = c(0, 3500), lwd = 3, 
     main = "Prior and Posterior for sigma2 (2001)", xlab = expression(sigma^2))
lines(sig2_vec, dinvgamma(sig2_vec, a, b), type = 'l', lwd = 3, col = 'red')
legend(x = 'topright', legend = c('Prior', 'Posterior'), col = c('red', 'black'),
       lty = 1, lwd = 3)

# code to plot the difference in posterior mean values (2022 - 2001)
plot(density(post.2022[keep, 1] - post.2001[keep, 1]), type = 'l',
     main = "Difference in Posterior Mean (2022 - 2001)", xlab = "Difference in Mean")

# code to test probability of difference in mean being less than 0
mean(post.2022[keep, 1] - post.2001[keep, 1] < 0)

# code to create 95% credible intervals for the difference in mean (2022 - 2001)
quantile(post.2022[keep, 1] - post.2001[keep, 1], c(0.025, 0.975))

# code to plot the ratio of variances (2022 / 2001)
plot(density(post.2022[keep, 2] / post.2001[keep, 2]), type = 'l',
     main = "Ratio of Variances (2022 / 2001)", xlab = expression(sigma^2))

# code to test probability of ratio of variances being greater than 1
mean((post.2022[keep, 2] / post.2001[keep, 2]) > 1)

# code to create 95% credible intervals for the ratio of variances (2022 / 2001)
quantile(post.2022[keep, 2] / post.2001[keep, 2], c(0.025, 0.975))

```
















